--- 
title: "DEQ R Methods Encyclopedia"
author: "DEQ R Development Team"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
# output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: openscapes/series
description: "Simple reprex of common tasks DEQ staff encounter and how to complete them with R."
split_by: "section"
---

# Background

The purpose of this book is to describe common tasks undertaken by DEQ staff and clearly outlines how to accomplish these objectives using the R programming language in order to promote learning opportunities for various skill levels of R users.

This project is an example of a [bookdown](https://bookdown.org/yihui/bookdown/) report built using R and [RStudio's R Markdown](http://rmarkdown.rstudio.com/). Many authors have contributed to this effort:

* Emma Jones (emma.jones@deq.virginia.gov)
* Connor Brogan (connor.brogan@deq.virginia.gov)
* Rex Robichaux (rex.robichaux@deq.virginia.gov)
* Joe Famularo (joseph.famularo@deq.virginia.gov)

Feel free to explore different chapters to learn about specific tasks and how to complete them efficiently and transparently in R. 

## Why R?{#whyR}

**Section Contact: ([joseph.famularo\@deq.virginia.gov](mailto:joseph.famularo@deq.virginia.gov))**

```{r child = 'subchapters/01-whyR.Rmd'}
```

## Getting Started With R{#gettingStartedWithR}

**Section Contact: Joe Famularo (joseph.famularo@deq.virginia.gov)**

```{r child = 'subchapters/01-DEQ_R_Introduction.Rmd'}
```

## Want to Contribute?{#wantToContribute}

We are always looking for collaborators. All [reprex](https://reprex.tidyverse.org/articles/articles/learn-reprex.html) (reproducible examples) must be published in R markdown using the bookdown package and bookdown output style. **Contributions are not limited to R as R markdown natively utilizes many languages including python and SQL.**

To get started with bookdown, see Yihui Xie's technical references book [bookdown: Authoring Books and Technical Documents with R Markdown](https://bookdown.org/yihui/bookdown/) or watch his webinar [introducing bookdown](https://www.rstudio.com/resources/webinars/introducing-bookdown/).

