# Query Internal Data Sources{#queryInternalDataSources}


```{r spatialAnalysis setup, include=FALSE}
```

Accessing internal DEQ data sources has been a longtime goal of many R programmers. Connecting a local R instance to raw data allows for increased automation and reduces redundant data copies on local drives. However, querying data directly from internal resources into one's local environment is a privilege and requires a number of R skills and approval from OIS. 

Database operations rely on the SQL language. Users must first familiarize themselves with the basics of database operations with SQL. The [DataCamp Introduction to SQL course](https://www.datacamp.com/courses/introduction-to-sql) is a good starting point. Additionally, the [DataCarpentry SQL databases and R](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html) course provides is a good practice before seeking a direct, read-only connection to ODS (the SQLServer back end of CEDS).

After you have **demonstrated proficiency** manipulating smaller, local data sources in R, you may submit a [Special Network Access Form](https://deqnet.deq.virginia.gov/programs/admin/ois/forms/documents/Special_Network_Access_Request_Form.pdf) to OIS seeking access to ODS data view using your Microsoft credentials. After OIS approval, you may access ODS from R following the connection instructions in the [Connect to ODS](#connectToODS) module. 

Alternatively, data "pinned" on the R server are available to all staff (on the internal network or VPN) without OIS authorization. Many pre-analyzed data sources are available to be pulled into your local R environment to assist with projects. See the [Connecting to R Connect (for Pinned Data)](#connectToConnectPins) module for more information on connecting your local environment to the R server to query these internal data resources. 

## Connecting to R Connect (for Pinned Data){#connectToConnectPins}

**Section Contact: Emma Jones (emma.jones@deq.virginia.gov)**

```{r child = 'subchapters/connectToConnectPins.Rmd'}
```

## Connect to ODS{#connectToODS}

**Section Contact: Emma Jones (emma.jones@deq.virginia.gov)**

Coming Soon!


## Querying Data from ODS{#queryDataFromODS}

Coming Soon!

## Common WQM queries{#commonWQMqueries}

Coming Soon!

## Common WQA queries{#commonWQAqueries}

Coming Soon!

## Common TMDL queries{#commonTMDLqueries}

Coming Soon!

