# Spatial Analysis {#spatialAnalysis}

```{r spatialAnalysis setup, include=FALSE}
```

R is a powerful tool for spatial analysis. Among the many benefits to using R as a GIS are the reproducibility of codified methods, which make sharing procedures and updating results after underlying data updates easy and efficient. 

There are many resources for learning how to use R for geospatial operations. A favorite is the [Geocomputation with R](https://geocompr.robinlovelace.net/) online book (free) by [Robin Lovelace](https://www.robinlovelace.net/).

To familiarize yourself with common geospatial procedures in R, the following chapter has been developed by DEQ staff to help colleagues with typical geospatial tasks. This is by no means a comprehensive introduction to geospatial techniques in R. Please see other resources for more background information on underlying geospatial principles and methods in R. 

## Spatial Data from Tabular Data{#spatialDataFromTabularData}

**Section Contact: Emma Jones (emma.jones@deq.virginia.gov)**


```{r child = 'subchapters/spatialDataFromTabularData.Rmd'}
```


## Consuming GIS REST Services in R{#consumingGISRESTservices}

**Section Contact:Rex Robichaux (rex.robichaux@deq.virginia.gov)**


```{r child = 'subchapters/Consuming_GIS_REST_Services_In_R.Rmd'}
```



## Using Spatial Datasets{#usingSpatialDatasets}

Coming Soon!

## Interactive Mapping{#interactiveMapping}

with leaflet. Coming soon!

## Watershed Delineation 

**Section Contact: Emma Jones (emma.jones@deq.virginia.gov)**

This module overviews the basics of delineating watersheds directly in your local R environment by scraping [USGS's StreamStats](https://www.usgs.gov/mission-areas/water-resources/science/streamstats-streamflow-statistics-and-spatial-analysis-tools) API. General web scraping techniques are beyond the scope of this module, but the basics can be gleaned by unpacking the referenced functions.

All watersheds delineated using this technique use USGS's StreamStats delineation techniques from a 1:24k NHD. You may manually explore the tool [here](https://streamstats.usgs.gov/ss/).


```{r child = 'subchapters/watershedDelineation.Rmd'}
```

## Landcover analysis

Coming Soon

## Spatial joins

Coming Soon

