# Spatial Analysis {#spatialAnalysis}

```{r spatialAnalysis setup, include=FALSE}
```


Coming soon: 

## Spatial Data from Tabular Data{#spatialDataFromTabularData}

Coming Soon!

## Using Spatial Datasets{#usingSpatialDatasets}

Coming Soon!

## Interactive Mapping{#interactiveMapping}

with leaflet. Coming soon!

## Watershed Delineation 

This module overviews the basics of delineating watersheds directly in your local R environment by scraping [USGS's StreamStats](https://www.usgs.gov/mission-areas/water-resources/science/streamstats-streamflow-statistics-and-spatial-analysis-tools) API. General web scraping techniques are beyond the scope of this module, but the basics can be gleaned by unpacking the referenced functions.

All watersheds delineated using this technique use USGS's StreamStats delineation techniques from a 1:24k NHD. You may manually explore the tool [here](https://streamstats.usgs.gov/ss/).


```{r child = 'subchapters/watershedDelineation.Rmd'}
```

## Landcover analysis

Coming Soon

## Spatial joins

Coming Soon

## Converting tabular data to spatial data

Coming Soon
