---
title: "Spatial Data from Tabular Data"
author: "Emma Jones"
date: "3/17/2022"
output: html_document
---

Location information can be critical to efficiently analyzing a dataset. Often, spatial data is included into spreadsheets as Latitude and Longitude fields, which can quickly be turned into spatial objects in R for further spatial analysis. This short example demonstrates a method of turning tabular data into spatial data, but this is not the only way to do so. 

```{r Spatial Data from Tabular Data setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, load in the necessary packages.

```{r Spatial Data from Tabular Data load packages}
library(tidyverse) # for tidy data manipulation
library(sf) # for spatial analysis
library(leaflet) # for interactive mapping
library(inlmisc) # for interactive mapping on top of leaflet
```

Create an example tabular dataset of stations that we want to make into a spatial dataset. Note: we are using the tibble::tribble() function to define a tibble row by row, which makes for easy visualization of test datasets. This is just one of many ways to create a [tibble](https://tibble.tidyverse.org/){target="_blank"}.

```{r Spatial Data from Tabular Data example data}
exampleSites <-  tribble(
  ~StationID, ~Latitude,  ~Longitude,
  "Station_1", 37.812840, -80.063946,
  "Station_2", 37.782322, -79.961449,
  "Station_3", 37.801644, -79.968441)

exampleSites
```

Using the [sf package](https://r-spatial.github.io/sf/){target="_blank"}, we can use the Latitude and Longitude fields to convert this object into a spatial object. Note we are using [EPSG 4326](https://epsg.io/4326){target="_blank"} for our [coordinate reference system](https://geocompr.robinlovelace.net/spatial-class.html#crs-intro){target="_blank"}. We chose that CRS 

```{r Spatial Data from Tabular Data st as sf}
exampleSites_sf <- exampleSites %>% 
  st_as_sf(coords = c("Longitude", "Latitude"),  # make spatial layer using these columns
           remove = F, # don't remove these lat/lon cols from df
           crs = 4326) # 
```

Now look at the difference in the object. The geometry listcolumn is where all the spatial magic is stored.

```{r Spatial Data from Tabular Data example sites sf}
exampleSites_sf
```

You can operate on a spatial dataset created using this method just like any other tidy object and the "sticky" geometry will come along for the ride.

```{r Spatial Data from Tabular Data sticky geometry}
justOneSite <- filter(exampleSites_sf, StationID == 'Station_1')
```

Let's plot the result to see what happened. 

```{r Spatial Data from Tabular Data map}
CreateWebMap(maps = c("Topo","Imagery","Hydrography"), collapsed = TRUE) %>%
  addCircleMarkers(data = exampleSites_sf, color='yellow', fillColor='black', radius = 5,
                     fillOpacity = 1, opacity=1,weight = 1,stroke=T, group="All Sites",
                     label = ~StationID,
                     popup=leafpop::popupTable(exampleSites_sf)) %>%
    addCircleMarkers(data = justOneSite, color='orange', fillColor='orange', radius = 5,
                     fillOpacity = 1, opacity=1,weight = 1,stroke=T, group="Just One Site",
                     label = ~StationID,
                     popup=leafpop::popupTable(justOneSite)) %>%
    addLayersControl(baseGroups=c("Topo","Imagery","Hydrography"),
                       overlayGroups = c('Just One Site', 'All Sites'),
                       options=layersControlOptions(collapsed=T),
                       position='topleft')
```


### Removing Spatial Information

Sometimes we need to remove spatial information in order to operate on a dataset (e.g. save just the tabular data as a csv). We can use sf to remove just the spatial information.

```{r Spatial Data from Tabular Data remove spatial info}
exampleSites_noSpatial <- exampleSites_sf %>% 
  st_drop_geometry()

exampleSites_noSpatial
```

